---
title: "ANCOVA with Combined Treatment Groups (Detailed)"
author: "junco"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ANCOVA with Combined Treatment Groups (Detailed)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

> **Short version available:**
> See *ANCOVA with Combined Treatment Groups (Short)* for a minimal working example
> with recommended defaults.

# Overview

This vignette provides a detailed explanation of **ANCOVA with combined treatment groups** in *junco*, including weighting strategies, interaction models, and alternative estimands.

# Clinical Context

ANCOVA is commonly used for covariate-adjusted analyses of change from baseline in clinical trials. Regulatory tables often require pooling of treatment arms while preserving the original estimand.

# Example Dataset

```{r}
library(dplyr)
library(rtables)
library(junco)

set.seed(101)

adchg <- tibble(
  USUBJID = sprintf("SUBJ-%03d", 1:300),
  TRT01A = factor(rep(c("Placebo", "Low Dose", "High Dose"), length.out = 300)),
  REGION = factor(sample(c("EU", "US"), 300, replace = TRUE, prob = c(0.6, 0.4))),
  BASE = rnorm(300, 50, 10),
  CHG  = rnorm(300, 0, 8)
)
```

# Combined Column Strategy

Combined treatment columns are defined using layout metadata and do not modify the underlying dataset.

```{r}
combodf <- tibble::tribble(
  ~valname,   ~label,    ~levelcombo,                     ~exargs,
  "ACTIVE",   "Active", c("Low Dose", "High Dose"), list()
)
```

# ANCOVA Model

The ANCOVA model is fitted as:

```
CHG ~ TRT01A + BASE + REGION
```

The model is fitted once using the original randomized treatment variable.

# Contrast-Based Combination (Default)

```{r}
lyt <- basic_table() %>%
  split_cols_by(
    "TRT01A",
    ref_group = "Placebo",
    split_fun = add_combo_levels(combodf)
  ) %>%
  add_colcounts() %>%
  analyze(
    vars = "CHG",
    afun = a_summarize_ancova_j,
    var_labels = "Change from Baseline",
    extra_args = list(
      variables = list(arm = "TRT01A", covariates = c("BASE", "REGION")),
      ref_path = c("TRT01A", "Placebo"),
      method_combo = "contrasts",
      weights_combo = "equal",
      weights_emmeans = "equal",
      conf_level = 0.95      
    )
  )

build_table(lyt, adchg)
```

# Weighting Strategies

- `equal`: equal contribution of arms
- `proportional`: weighted by observed sample size
- `proportional_marginal`: marginal weights across interaction strata

# Interaction Models

If an interaction such as `TRT01A * REGION` is included, combined estimates are derived within strata, and weighting determines how pooling is applied.

# Statistical Appendix: emmeans-based Estimation

## Notation

Let:

- \( \hat{\mu}_k \) denote the LS mean for treatment arm \( k \)
- \( \mathbf{\hat{\mu}} = (\hat{\mu}_1, \ldots, \hat{\mu}_K)^T \)
- \( \mathbf{V} = \mathrm{Var}(\mathbf{\hat{\mu}}) \)

All quantities are obtained from the fitted ANCOVA model.

---

## A. Combined LS Means

A combined treatment LS mean for a set of arms \( C \) is defined as:

\[
\hat{\mu}_{C} = \sum_{k \in C} w_k \, \hat{\mu}_k ,
\qquad \sum_{k \in C} w_k = 1
\]

where \( w_k \) are user-specified weights (e.g. equal or proportional).

### Variance

The variance of the combined LS mean is:

\[
\mathrm{Var}(\hat{\mu}_{C}) = \mathbf{w}^T \, \mathbf{V} \, \mathbf{w}
\]

This corresponds exactly to a linear contrast implemented via
`emmeans::contrast()`.

---

## B. Contrasts Between Treatment Groups

### Simple Treatment Differences

The estimated difference between two individual treatment arms \( k \) and \( r \)
(e.g. Active vs Placebo) is defined as the contrast:

\[
\hat{\delta}_{k,r} = \hat{\mu}_k - \hat{\mu}_r
\]

with variance:

\[
\mathrm{Var}(\hat{\delta}_{k,r}) = \mathrm{Var}(\hat{\mu}_k)
+ \mathrm{Var}(\hat{\mu}_r)
- 2 \, \mathrm{Cov}(\hat{\mu}_k, \hat{\mu}_r)
\]

This covariance term is automatically accounted for when contrasts are derived
from the joint covariance matrix \( \mathbf{V} \).

---

### Differences Involving Combined Arms

For a combined treatment \( C \) compared to reference arm \( r \):

\[
\hat{\delta}_{C,r} = \hat{\mu}_{C} - \hat{\mu}_r
\]

This can be written as a single linear contrast with weight vector:

\[
\mathbf{w}_{C,r} = (w_1, \ldots, w_K, -1)
\]

and variance:

\[
\mathrm{Var}(\hat{\delta}_{C,r}) = \mathbf{w}_{C,r}^T \, \mathbf{V} \, \mathbf{w}_{C,r}
\]

This ensures internally consistent standard errors, confidence intervals, and
p-values for combined treatment comparisons.

---

## C. Contrasts in Interaction Models

When the ANCOVA model includes interaction terms (e.g. `TRT01A * REGION`),
LS means are estimated **within each interaction stratum**.

In this setting:

- Treatment-specific LS means \( \hat{\mu}_{k,s} \) are defined per stratum \( s \)
- Combined means are first formed *within stratum*
- Pooling across strata is then applied via user-defined weights

Formally, the combined LS mean marginal over strata is:

\[
\hat{\mu}_{C} = \sum_{s} \pi_s \,
\Big( \sum_{k \in C} w_{k|s} \, \hat{\mu}_{k,s} \Big)
\]

where:

- \( w_{k|s} \) are within-stratum treatment weights
- \( \pi_s \) are stratum-level weights

The distinction between `weights_combo = "proportional"` and
`"proportional_marginal"` determines whether \( w_{k|s} \) or \( \pi_s \)
reflect observed sample sizes within or across strata.

This two-stage weighting clarifies how estimands change in the presence of
interactions and avoids implicit, undocumented averaging.

---

# Collapsing Treatment Arms

```r
method_combo = "collapse"
```

This refits ANCOVA on pooled arms and changes the estimand.
Use only if explicitly specified in the SAP.

# Summary

- Combined LS means and differences are implemented as linear contrasts
- Full covariance propagation is preserved
- Interaction models require explicit decisions on within- and across-stratum weighting
- The approach is fully aligned with `emmeans` theory
