---
title: "ANCOVA with Combined Treatment Groups (Short)"
author: "junco"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ANCOVA with Combined Treatment Groups (Short)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

> 📘 **Looking for more detail?**
> See *ANCOVA with Combined Treatment Groups (Detailed)* for weighting strategies,
> interaction models, and the mathematical definition of combined LS means.

# Purpose

This short vignette demonstrates how **junco** supports ANCOVA tables with **combined treatment columns**, while fitting the model on the original randomized treatment groups.

# Key Principle

> **Estimation and presentation are separated**.

- ANCOVA is fitted once using the original treatment variable.
- Combined arms are created only at the table layout level.

# Clinical Example

```{r}
library(dplyr)
library(rtables)
library(junco)

set.seed(101)

adchg <- tibble(
  USUBJID = sprintf("SUBJ-%03d", 1:300),
  TRT01A = factor(rep(c("Placebo", "Low Dose", "High Dose"), length.out = 300)),
  REGION = factor(sample(c("EU", "US"), 300, replace = TRUE)),
  BASE = rnorm(300, 50, 10),
  CHG  = rnorm(300, 0, 8)
)
```

# Combined Column Definition

```{r}
combodf <- tibble::tribble(
  ~valname, ~label,   ~levelcombo,                     ~exargs,
  "ACTIVE", "Active", c("Low Dose", "High Dose"), list()
)
```

# ANCOVA Table

```{r}
lyt <- basic_table() %>%
  split_cols_by(
    "TRT01A",
    ref_group = "Placebo",
    split_fun = add_combo_levels(combodf)
  ) %>%
  analyze(
    vars = "CHG",
    afun = a_summarize_ancova_j,
    extra_args = list(
      variables = list(arm = "TRT01A", covariates = c("BASE", "REGION")),
      ref_path = c("TRT01A", "Placebo"),
      method_combo = "contrasts",
      weights_combo = "equal",
      weights_emmeans = "equal",
      conf_level = 0.95
    )
  )

build_table(lyt, adchg)
```

# Summary

- Fit ANCOVA on original arms
- Combine treatments via contrasts
- Default and recommended: `method_combo = "contrasts"`
