---
title: "Formatting function for significant figures"
date: "`r Sys.Date()`"
output:
    rmarkdown::html_document:
        theme: "spacelab"
        highlight: "kate"
        toc: true
        toc_float: true
vignette: >
  %\VignetteIndexEntry{Formatting function for significant figures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
      wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

Many clinical tables require statistics presented on a table with certain decimal precision.
In some situations, the usage of significant figures is preferred over a fixed decimal precision. A typical situation is in the context of PK tables.

The purpose in this vignette is to provide some examples for `format_sigfig_j` and the difference between `tern::format_sigfig()`.

```{r load_packages, message=FALSE}
library(rtables)
library(junco)
library(dplyr)
library(tern)
```


## Significant Figures and Company Specific Requirements

For PK tables, the requirements within our company to present numerical values of core PK parameters with 3 significant figures, except for values above 1000, the whole integer has to be reported.
E.g. 3 significant figures for a value 1234.567 is `signif(1234.567, 3)` = `r signif(1234.567, 3)`, while a value of `r round_fmt(1234.567, 0)` is expected.

In addition, we want to adhere to possibility to adjust the type of rounding to be `SAS`, `iec` or `iec_mod`, through the `formatters` functions `format_value` and `round_fmt`.
The base function `sigfig` relies on `round`, so only supports `iec` rounding type. 

For these 2 reasons, the formatting function factory `format_sigfig_j` has been developed as alternative to `tern::format_sigfig`.

## Basic Examples of Formatting Function Factory `format_sigfig_j`
Below are a few examples where the difference in behavior between `signif`, `tern::format_sigfig` and `junco::format_sigfig_j` is demonstrated.

```{r example}
fmt <- format_sigfig_j(3)
fmt(0.001234)
fmt(1234.5)
fmt(1234.5, round_type = "sas")
signif(1234.5, 3)
format_sigfig(3)(1234.5)

fmt2 <- format_sigfig_j(3, format = "xx (xx)")
fmt2(c(1234.5, 0.00567))

format_sigfig_j(3, drop0trailing = FALSE)(0.01)
format_sigfig_j(3, drop0trailing = TRUE)(0.01)
format_sigfig(3)(0.01)
```
## Formats based upon `junco::format_sigfig_j` in `rtables` Layout

### Example as in Tern TLG Catalog using `analyze_vars_in_cols`
The following example is taken from the [**tern TLG catalog**](https://insightsengineering.github.io/tlg-catalog/stable/tables/pharmacokinetic/pkct01.html).

Here is the data that is being used. 
```{r example 2 - datasetup}
adsl <- random.cdisc.data::cadsl |>
  filter(ACTARM == "A: Drug X")
adpc <- random.cdisc.data::cadpc |>
  filter(ACTARM == "A: Drug X", PARAM %in% c("Plasma Drug X", "Urine Drug X"))

# Setting up the data
adpc_1 <- adpc |>
  mutate(
    NFRLT = as.factor(NFRLT),
    AVALCAT1 = as.factor(AVALCAT1),
    NOMTPT = as.factor(paste(NFRLT, "/", PCTPT))
  ) |>
  select(STUDYID, NOMTPT, ACTARM, VISIT, AVAL, PARAM, AVALCAT1)

adpc_1$NOMTPT <- factor(
  adpc_1$NOMTPT,
  levels = levels(adpc_1$NOMTPT)[order(as.numeric(gsub(".*?([0-9\\.]+).*", "\\1", levels(adpc_1$NOMTPT))))]
)
```

The following row-structure setup in the layout definition is utilized.
```{r example 2 - layout with analyze_vars_in_cols}
# Row structure
lyt_rows <- basic_table() |>
  split_rows_by(
    var = "PARAM",
    split_fun = drop_split_levels,
    split_label = "Parameter",
    label_pos = "topleft"
  ) |>      
  split_rows_by(
    var = "ACTARM",
    split_fun = drop_split_levels,
    split_label = "Treatment Group",
    label_pos = "topleft"
  ) |>
  split_rows_by(
    var = "VISIT",
    split_fun = drop_split_levels,
    split_label = "Visit",
    label_pos = "topleft"
  ) |>
  split_rows_by(
    var = "NOMTPT",
    split_fun = drop_split_levels,
    split_label = "Nominal Time (hr) / Timepoint",
    label_pos = "topleft",
    child_labels = "hidden"
  )
```

The core component of the layout where the significant figures format specifications are used is in below call to `tern::analyze_vars_in_cols`.
The only adjustment made here compared to the example in the **tern TLG catalog** is the usage of formatting function `junco::format_sigfig_j(3)` versus `tern::format_sigfig(3)`.

```{r example 2 - continue layout}
lyt <- lyt_rows |>
  analyze_vars_in_cols(
    vars = c("AVAL", "AVALCAT1", rep("AVAL", 8)),
    .stats = c("n", "n_blq", "mean", "sd", "cv", "geom_mean", "geom_cv", "median", "min", "max"),
    .formats = c(
      n = "xx.", n_blq = "xx.", mean = format_sigfig_j(3), sd = format_sigfig_j(3), cv = "xx.x", median = format_sigfig_j(3),
      geom_mean = format_sigfig_j(3), geom_cv = "xx.x", min = format_sigfig_j(3), max = format_sigfig_j(3)
    ),
    .labels = c(
      n = "n", n_blq = "Number\nof\nLTRs/BLQs", mean = "Mean", sd = "SD", cv = "CV (%) Mean",
      geom_mean = "Geometric Mean", geom_cv = "CV % Geometric Mean", median = "Median", min = "Minimum", max = "Maximum"
    ),
    na_str = "NE",
    .aligns = "decimal"
  )

result <- build_table(lyt, df = adpc_1, alt_counts_df = adsl) 

result
```

### Same Example with `junco` afun `column_stats`

An alternative, very similar layout can be generated using the `junco` afun `column_stats`.
The advantage of this afun, is that it can deal with so called, varying decimal precision.

In case, not all parameters of the table are to be presented with the same number of significant figures, the following example demonstrates how this can be achieved using `junco::column_stats`.

``` {r setup - stats_in_cols}
 mysplitfun <- make_split_fun(
   post = list(stats_in_cols_setup())
 )

 lyt_stats_in_cols <- basic_table() |>
   split_cols_by_multivar(
     c("AVAL"),
     varlabels = c("Visit Values")
   ) |>
   split_cols_by("STUDYID", split_fun = mysplitfun) |>
   split_rows_by("PARAM", split_fun = drop_split_levels) |>
   split_rows_by("ACTARM", labels_var = "ACTARM", split_fun = drop_split_levels) |>
   split_rows_by("VISIT", split_fun = drop_split_levels) 
 
 
 adpc_1 <- adpc_1 |> 
   filter(VISIT == "Day 1", !(NOMTPT %in% "0 / Predose")) 
```

Here 2 methods of setting up format specification based upon `format_sigfig_j` are presented.

The first one is a manual setup for each parameter separately.

``` {r setup - format 1}
 adpc_fmt <- tribble(
   ~PARAM, ~fmt_d,
   "Plasma Drug X", list(
     mean = format_sigfig_j(3), sd = format_sigfig_j(3), se = format_sigfig_j(3), median = format_sigfig_j(3),
     min = format_sigfig_j(3), max = format_sigfig_j(3)
   ),
   "Urine Drug X", list(
     mean = format_sigfig_j(2), sd = format_sigfig_j(2), se = format_sigfig_j(2), median = format_sigfig_j(2),
     min = format_sigfig_j(2), max = format_sigfig_j(2)
   )
 )

adpc_2 <- left_join(adpc_1, adpc_fmt) 
```

The second one is a setup, using the same tools as in [vignette on Formatting tools to simplify adjustment of decimal precision](vignette("formats_d", package = "junco") (`fmt_spec_df_d`), to convert a simple dataframe setup into a dataframe with formatting specifications.

``` {r setup - format 2}
junco_def_d_sigfig <- list(
     mean = format_sigfig_j(3), sd = format_sigfig_j(3), se = format_sigfig_j(3), median = format_sigfig_j(3),
     min = format_sigfig_j(3), max = format_sigfig_j(3)
   )

adpc_fmt2 <- tribble(
   ~PARAM, ~d,
   "Plasma Drug X", 3,
   "Urine Drug X", 2
 )

adpc_fmt2 <- fmt_spec_df_d(adpc_fmt2,
                    d_column = "d",
                    fmt_column = "fmt_d2",
                    fmt_d_def = junco_def_d_sigfig,
                    fmt_d_in = NULL
)


fmt_details <- lapply(adpc_fmt2$PARAM,
                      FUN = function(param){
                        fmt <- adpc_fmt2[adpc_fmt2$PARAM == param, c("fmt_d2"), drop = TRUE][[1]]
                        fmt
                        get_fmt_details(fmt, recursive = FALSE, as_tibble = TRUE)
                      }
  )

names(fmt_details) <- adpc_fmt2$PARAM

# confirm 2 significant figures for Urine Drug X
# see column d in tibble
print(fmt_details[["Urine Drug X"]][, c("stat", "fun_fact", "str", "d", "drop0trailing", "whole_integer")])

adpc_3 <- left_join(adpc_1, adpc_fmt2) 
```

Now, both layouts are being finalized and results are created. 
```{r example 2 - layout 2}
 lyt2 <- lyt_stats_in_cols |> 
   analyze("NOMTPT", afun = column_stats, extra_args = list(
     .formats = "default",
     formats_var = "fmt_d"
   ))
 result2 <- build_table(lyt2, adpc_2, round_type = "sas") |> 
   prune_table()
 result2
```

```{r example 2 - layout 3}
lyt3 <- lyt_stats_in_cols |> 
 analyze("NOMTPT", afun = column_stats, extra_args = list(
   .formats = "default",
   formats_var = "fmt_d2"
 ))
result3 <- build_table(lyt3, adpc_3, round_type = "sas") |> 
 prune_table()
result3
```
Confirmation that both approaches yield the same results (raw + formatted values).
```{r example 2 - identical result}
identical(cell_values(result2), cell_values(result3))
identical(get_formatted_cells(result2), get_formatted_cells(result3))
```