---
title: "Reference Group Handling with get_ref_info"
date: "`r Sys.Date()`"
author: "David Munoz Tord"
output:
    rmarkdown::html_document:
        theme: "spacelab"
        highlight: "kate"
        toc: true
        toc_float: true
vignette: >
  %\VignetteIndexEntry{Reference Group Handling with get_ref_info}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
      wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

Many clinical tables require statistics computed relative to a reference
(control) group, for example, a difference in means versus placebo, or
a risk difference versus a comparator arm. In `rtables`, the reference
group is normally injected automatically via `.ref_group` and
`.in_ref_col` when the analysis function is placed inside the reference
column's split. This works well for simple single-level column splits,
but breaks down when:

- the reference column is nested under a spanning header (e.g. a
  `colspan_trt` variable),
- a `split_cols_by_multivar` is present, or
- an `add_overall_col` is used alongside treatment splits.

`get_ref_info()` solves this by letting the analysis function look up
the reference group itself, using an explicit column path (`ref_path`).

```{r load_packages, message=FALSE}
library(rtables)
library(junco)
library(tern)
library(dplyr)
```

```{r load_data, message=FALSE}
adsl <- ex_adsl
```

## The `ref_path` Convention

`ref_path` is a character vector of alternating split-variable names and
their values, following the same convention as `rtables::col_paths()`:

```
c("var1", "level1", "var2", "level2", ...)
```

For a simple layout with `split_cols_by("ARM")`, the placebo reference
path is:

```r
ref_path <- c("ARM", "B: Placebo")
```

For a layout with a spanning header variable `colspan_trt` above `ARM`,
the path must include both levels:

```r
ref_path <- c("colspan_trt", " ", "ARM", "B: Placebo")
```

The path must exactly match the column-split hierarchy in the layout.
Use `col_paths(build_table(lyt, df))` to inspect the available paths
for a given layout.

## Basic Usage

`get_ref_info()` returns a list with two elements:

- `in_ref_col`: `TRUE` if the current column matches `ref_path`,
  `FALSE` otherwise, `NULL` if the reference split variable is not
  present in the current column hierarchy.
- `ref_group`: the subset of the row data belonging to the reference
  group (a data frame, or a vector if `.var` is supplied).

Here is a minimal analysis function that uses `get_ref_info()`:

```{r basic_afun}
afun_diff_means <- function(x, .var, ref_path, .spl_context) {
  ref <- get_ref_info(ref_path, .spl_context, .var)

  diff <- if (isFALSE(ref$in_ref_col)) {
    mean(x, na.rm = TRUE) - mean(ref$ref_group, na.rm = TRUE)
  } else {
    NULL
  }

  in_rows(
    m    = rcell(mean(x, na.rm = TRUE), label = "Mean", format = "xx.x"),
    diff = rcell(diff, label = "Diff vs Placebo", format = "xx.x")
  )
}
```

Applied to a simple single-level column split:

```{r basic_table}
ref_path <- c("ARM", "B: Placebo")

lyt <- basic_table() |>
  split_cols_by("ARM") |>
  analyze("AGE", afun = afun_diff_means, extra_args = list(ref_path = ref_path))

build_table(lyt, adsl)
```

## With a Spanning Header

When a spanning header variable (`colspan_trt`) sits above the treatment
split, the reference column is nested one level deeper. The `ref_path`
must include the spanning variable:

```{r colspan_setup}
adsl2 <- adsl |>
  mutate(
    colspan_trt = factor(
      ifelse(ARM == "B: Placebo", " ", "Active Study Agent"),
      levels = c("Active Study Agent", " ")
    )
  )

colspan_trt_map <- create_colspan_map(
  adsl2,
  non_active_grp         = "B: Placebo",
  non_active_grp_span_lbl = " ",
  active_grp_span_lbl    = "Active Study Agent",
  colspan_var            = "colspan_trt",
  trt_var                = "ARM"
)

ref_path_nested <- c("colspan_trt", " ", "ARM", "B: Placebo")
```

```{r colspan_table}
lyt2 <- basic_table() |>
  split_cols_by("colspan_trt", split_fun = trim_levels_to_map(colspan_trt_map)) |>
  split_cols_by("ARM") |>
  analyze("AGE", afun = afun_diff_means, extra_args = list(ref_path = ref_path_nested))

build_table(lyt2, adsl2)
```

Note that `get_ref_info()` returns `in_ref_col = NULL` (not `FALSE`)
when the reference split variable is absent from the current column
hierarchy — for example, in an overall column added via
`add_overall_col()`. The analysis function should handle `NULL`
explicitly:

```{r overall_col}
lyt3 <- basic_table() |>
  split_cols_by("colspan_trt", split_fun = trim_levels_to_map(colspan_trt_map)) |>
  split_cols_by("ARM") |>
  add_overall_col("Total") |>
  analyze("AGE", afun = afun_diff_means, extra_args = list(ref_path = ref_path_nested))

build_table(lyt3, adsl2)
```

The `Total` column receives `in_ref_col = NULL` because `colspan_trt`
is not part of its column path. The analysis function returns `NULL` for
the difference row, producing a blank cell.

## Wildcard Matching

`in_column()` — the underlying helper used by `get_ref_info()` — supports
`"*"` as a wildcard for any single split variable name or value. This is
useful when you want to check whether the current column belongs to a
particular split level regardless of the spanning header above it:

```{r wildcard}
# Matches any column where ARM == "B: Placebo", regardless of spanning header
spl_context_placebo <- data.frame(
  cur_col_split     = I(list(c("colspan_trt", "ARM"))),
  cur_col_split_val = I(list(c(" ", "B: Placebo")))
)

# exact match
in_column(c("colspan_trt", " ", "ARM", "B: Placebo"), spl_context_placebo)

# wildcard on the spanning header — still TRUE
in_column(c("*", "*", "ARM", "B: Placebo"), spl_context_placebo)

# wildcard does not match a different ARM level
in_column(c("*", "*", "ARM", "A: Drug X"), spl_context_placebo)
```

## Relationship to `h_get_trtvar_refpath`

`get_ref_info()` and `h_get_trtvar_refpath()` serve different purposes
and are used in different contexts:

| | `get_ref_info()` | `h_get_trtvar_refpath()` |
|---|---|---|
| **Returns** | reference-group data + `in_ref_col` flag | treatment variable name, current group, reference group name |
| **Used for** | obtaining `.ref_group` / `.in_ref_col` for statistics | identifying the treatment variable when `split_cols_by_multivar` is present |
| **Typical callers** | `s_*` / `a_*` analysis functions | `a_summarize_aval_chg_diff_j`, `a_freq_j`, `a_eair_j` |

In layouts that combine `split_cols_by_multivar` with difference
columns, `h_get_trtvar_refpath()` is called first to resolve the
treatment variable name from the interleaved column path, and
`get_ref_info()` is then called to obtain the reference-group data.

## Inspecting Column Paths

When constructing `ref_path`, it is helpful to inspect the column paths
of a built table:

```{r col_paths}
tbl <- build_table(lyt2, adsl2)
col_paths(tbl)
```

Each path in the output corresponds to a leaf column. The `ref_path`
should match one of these paths exactly (or use `"*"` wildcards for
positions that should not be constrained).
```
