---
title: "Reference Group Handling with get_ref_info"
date: "`r Sys.Date()`"
output:
    rmarkdown::html_document:
        theme: "spacelab"
        highlight: "kate"
        toc: true
        toc_float: true
vignette: >
  %\VignetteIndexEntry{Reference Group Handling with get_ref_info}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
      wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

Many clinical tables require statistics computed relative to a reference
(control) group — for example, a risk difference versus placebo. In
`rtables`, the reference group can be injected automatically via
`.ref_group` and `.in_ref_col`, but this breaks down when:
 
- the reference column is nested under a spanning header,
- a `split_cols_by_multivar` is present, or
- an `add_overall_col` is used alongside treatment splits.

`get_ref_info()` solves this by letting the analysis function look up
the reference group itself, using an explicit column path (`ref_path`).

```{r load_packages, message=FALSE}
library(rtables)
library(junco)
library(dplyr)
```

## The `ref_path` Convention

`ref_path` is a character vector of alternating split-variable names and
their values, following the same convention as `rtables::col_paths()`:

```
c("var1", "level1", "var2", "level2", ...)
```

For a simple layout with `split_cols_by("ARM")`, the placebo reference
path is:

```r
ref_path <- c("ARM", "B: Placebo")
```

For a layout with a spanning header variable `colspan_trt` above `ARM`,
the path must include both levels:

```r
ref_path <- c("colspan_trt", " ", "ARM", "B: Placebo")
```

The path must exactly match the column-split hierarchy in the layout.
Use `col_paths(build_table(lyt, df))` to inspect the available paths.

## Basic Usage

`get_ref_info()` returns a list with two elements:

- `in_ref_col`: `TRUE` if the current column matches `ref_path`,
  `FALSE` otherwise, `NULL` if the reference split variable is not
  present in the current column hierarchy.
- `ref_group`: the subset of the row data belonging to the reference
  group (a data frame, or a vector if `.var` is supplied).

## Working Example: AE Table with Risk Difference Columns

This example demonstrates the standard pattern for a table with
spanning headers and risk difference columns — the primary use case
for `get_ref_info()`.

```{r data_prep}
trtvar <- "TRT01A"
ctrl_grp <- "Placebo"

adsl <- pharmaverseadamjnj::adsl |>
  filter(SAFFL == "Y") |>
  select(STUDYID, USUBJID, all_of(trtvar), SAFFL) |>
  mutate(!!trtvar := factor(
    .data[[trtvar]],
    levels = c("Xanomeline Low Dose", "Xanomeline High Dose", "Placebo")
  ))

adae <- pharmaverseadamjnj::adae |>
  filter(TRTEMFL == "Y") |>
  select(USUBJID, TRTEMFL, AEBODSYS, AEDECOD)

# Add spanning header and risk difference variables
adsl$colspan_trt <- factor(
  ifelse(adsl[[trtvar]] == ctrl_grp, " ", "Active Study Agent"),
  levels = c("Active Study Agent", " ")
)
adsl$rrisk_header <- "Risk Difference (%) (95% CI)"
adsl$rrisk_label <- paste(adsl[[trtvar]], "vs", ctrl_grp)

ae <- adae |> right_join(adsl, by = "USUBJID")
```

The `ref_path` must trace the full column-split hierarchy down to the
reference group:

```{r ref_path}
colspan_trt_map <- create_colspan_map(
  adsl,
  non_active_grp = ctrl_grp,
  non_active_grp_span_lbl = " ",
  active_grp_span_lbl = "Active Study Agent",
  colspan_var = "colspan_trt",
  trt_var = trtvar
)

ref_path <- c("colspan_trt", " ", trtvar, ctrl_grp)
```

Now build the layout. The key point: `a_freq_j` uses `get_ref_info()`
internally to identify the reference group and compute risk differences
in the dedicated difference columns:

```{r layout_and_table}
extra_args <- list(
  denom = "n_altdf",
  riskdiff = TRUE,
  ref_path = ref_path,
  method = "wald",
  .stats = "count_unique_fraction",
  .formats = c(rr_ci_3d = jjcsformat_xx("xx.x (xx.x, xx.x)"))
)

lyt <- basic_table(show_colcounts = TRUE, colcount_format = "N=xx") |>
  split_cols_by("colspan_trt",
    split_fun = trim_levels_to_map(map = colspan_trt_map)
  ) |>
  split_cols_by(trtvar) |>
  split_cols_by("rrisk_header", nested = FALSE) |>
  split_cols_by(trtvar,
    labels_var = "rrisk_label",
    split_fun = remove_split_levels(ctrl_grp)
  ) |>
  analyze("TRTEMFL",
    afun = a_freq_j,
    extra_args = append(extra_args, list(val = "Y", label = "Subjects with >=1 AE"))
  ) |>
  split_rows_by("AEBODSYS",
    split_label = "System Organ Class",
    split_fun = trim_levels_in_group("AEDECOD"),
    label_pos = "topleft",
    section_div = " ",
    nested = FALSE
  ) |>
  summarize_row_groups("AEBODSYS", cfun = a_freq_j, extra_args = extra_args) |>
  analyze("AEDECOD", afun = a_freq_j, extra_args = extra_args) |>
  append_topleft("  Preferred Term, n (%)")

result <- build_table(lyt, ae, alt_counts_df = adsl)
head(result, 10)
```

## How `get_ref_info` Works Internally

When `a_freq_j` (or any analysis function) calls
`get_ref_info(ref_path, .spl_context)`:
 
1. It checks whether the split variables in `ref_path` are present in
   the current column hierarchy (using `in_column()` with wildcards).
2. If present, it subsets the row data to the reference group using the
   pre-computed column facet indices stored in `.spl_context`.
3. It returns `in_ref_col = TRUE/FALSE` indicating whether the current
   column IS the reference column (so the function can skip computing
   a difference against itself).

When the reference split is absent from the current column hierarchy
(e.g. in an `add_overall_col("Total")` column), both `in_ref_col` and
`ref_group` are `NULL`.

## Wildcard Matching with `in_column()`

`in_column()` — the underlying helper used by `get_ref_info()` —
supports `"*"` as a wildcard for any single split variable name or
value:

```{r wildcard}
# Simulate a split context for ARM == "B: Placebo" under colspan_trt == " "
spl_ctx <- data.frame(
  cur_col_split     = I(list(c("colspan_trt", "ARM"))),
  cur_col_split_val = I(list(c(" ", "B: Placebo")))
)

# Exact match
in_column(c("colspan_trt", " ", "ARM", "B: Placebo"), spl_ctx)

# Wildcard on spanning header — still TRUE
in_column(c("*", "*", "ARM", "B: Placebo"), spl_ctx)

# Different ARM level — FALSE
in_column(c("*", "*", "ARM", "A: Drug X"), spl_ctx)
```

## Relationship to `h_get_trtvar_refpath`

`get_ref_info()` and `h_get_trtvar_refpath()` serve different purposes:

| | `get_ref_info()` | `h_get_trtvar_refpath()` |
|---|---|---|
| **Returns** | reference-group data + `in_ref_col` flag | treatment variable name, current group, reference group name |
| **Used for** | obtaining `.ref_group` / `.in_ref_col` for statistics | identifying the treatment variable when `split_cols_by_multivar` is present |
| **Typical callers** | `a_freq_j`, `a_eair_j`, `s_ancova_j` | `a_summarize_aval_chg_diff_j` |

In layouts that combine `split_cols_by_multivar` with difference
columns, `h_get_trtvar_refpath()` resolves the treatment variable name
from the interleaved column path, and `get_ref_info()` obtains the
reference-group data.

## Inspecting Column Paths

When constructing `ref_path`, inspect the column paths of a built table:

```{r col_paths}
col_paths(result)
```

Each path corresponds to a leaf column. The `ref_path` should match one
of these paths exactly (or a prefix thereof when the reference column
has further nested splits below it).
